---
title: "Incidence Pipeline: From Microdata to Plot"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
pkgload::load_all(quiet = TRUE)
library(dplyr)
```

## From Many Policies to One Comparison

A fiscal microsimulation produces multiple datasets — a baseline and one or more
reforms. Each scenario yields the same household survey structure but with
different values (because the policy changed who pays what and who receives
what). The incidence pipeline takes these separate datasets and produces a
single plot-ready tibble that enables direct comparison across policies and
income groups.

### Three Phases

The pipeline operates in three conceptual phases:

| Phase | What happens | Data shape |
|-------|-------------|------------|
| **Separate** | Each simulation is processed independently. Deciles are computed per-policy (a reform might change a household's income rank). | List of data frames |
| **Merge** | All simulations are aggregated into summary statistics and stacked into one tibble. A `sim` column tags which policy each row belongs to. | One long tibble |
| **Transform** | Incidence measures are calculated, labels applied, and columns renamed — all uniformly across the combined data. | One plot-ready tibble |

### Why This Order

- Deciles **must** be computed per-simulation (a reform changes income rankings)
- Aggregation is where we lose household-level detail and gain comparability
- After merging, every operation treats all policies identically — `sim` is
  just another grouping variable

### Pipeline Diagram

```{mermaid}
flowchart LR
  subgraph SEPARATE["Phase 1 · Separate"]
    direction TB
    A1[Baseline\npolicy_sim_raw]
    A2[Reform\npolicy_sim_raw]
    B[f_calc_deciles_by_sim]
    A1 --> B
    A2 --> B
    B --> C1[Baseline\n+ decile cols]
    B --> C2[Reform\n+ decile cols]
  end

  subgraph MERGE["Phase 2 · Merge"]
    direction TB
    D[f_agg_by_decile_by_sim]
    E[One tibble\nsim · decile · var · level\ngroup_var · group_val · pop]
    D --> E
  end

  subgraph TRANSFORM["Phase 3 · Transform"]
    direction TB
    F[f_calc_incidence]
    G[f_format_incidence]
    H[Plot-ready tibble\nSimulation · Decile · Variable\nStatistics · Value · Group]
    F --> G --> H
  end

  C1 --> D
  C2 --> D
  H --> I[f_plot_gg\nx=Decile · color=Simulation]

  style SEPARATE fill:#e8f4fd,stroke:#2196F3
  style MERGE fill:#fff3e0,stroke:#FF9800
  style TRANSFORM fill:#e8f5e9,stroke:#4CAF50
```

---

## The Input: Simulation Data

The pipeline starts with `dta_sim` — a named list where each element represents
one policy scenario.

```{r input-structure}
# Named list of policy simulations
names(dta_sim)

# Each element: list(policy_sim_raw = <tibble>, policy_name = <character>)
str(dta_sim[[1]][["policy_name"]])
str(dta_sim[[2]][["policy_name"]])
```

Each `policy_sim_raw` is a household-level data frame with income concepts,
fiscal variables, weights, and grouping variables:

```{r input-columns}
dta_sim[[1]][["policy_sim_raw"]] |>
  select(hhid, hhwt, ym, yn, dtx_prog1, dtr_prog1, group_1) |>
  head(5)
```

---

## Phase 1: Separate — Assign Deciles

### What `f_calc_deciles()` does (single dataset)

Before seeing the full simulation wrapper, here is how decile assignment works
on a single data frame. Given income variables and a weight, it creates factor
columns ranking each household from 1 (poorest) to N (richest):

```{r single-deciles}
single_df <- dta_sim[[1]][["policy_sim_raw"]]

single_with_dec <- single_df |>
  f_calc_deciles(dec_var = c("ym", "yn"), wt_var = "hhwt", n_dec = 10)

single_with_dec |>
  select(hhid, ym, ym___decile, yn, yn___decile) |>
  head(5)
```

The `___decile` suffix is the naming convention. Each column is an ordered
factor (1–10).

### Applying across simulations: `f_calc_deciles_by_sim()`

This wrapper applies `f_calc_deciles()` to each simulation independently. The
output is the **same list structure** — just with decile columns added:

```{r step1-deciles}
sim_with_deciles <- dta_sim |>
  f_calc_deciles_by_sim(
    dec_var = c("ym", "yn"),
    wt_var = "hhwt",
    n_dec = 10
  )

# Still a list, same length
length(sim_with_deciles)

# Each element still has policy_sim_raw + policy_name
names(sim_with_deciles[[1]])

# But now policy_sim_raw has the decile columns
sim_with_deciles[[2]][["policy_sim_raw"]] |>
  select(hhid, ym, ym___decile) |>
  head(3)
```

Deciles are computed independently per simulation because a policy reform can
change a household's income and therefore its rank.

---

## Phase 2: Merge — Aggregate and Stack

### What `f_agg_by_decile_one()` does (single variable)

The core aggregation function computes weighted sums of a variable within each
decile. Here is how it works on a single dataset, one decile variable, and one
aggregation variable:

```{r single-agg-one}
single_with_dec |>
  f_agg_by_decile_one(
    var_decile = "ym___decile",
    var_agg = "dtx_prog1",
    var_group = "group_1",
    wt_var = "hhwt"
  ) |>
  head(8)
```

Output columns:

- `decile_var` / `decile` — which decile grouping and which decile
- `var` / `level` — which variable was aggregated and the weighted sum
- `group_var` / `group_val` — disaggregation dimension and category
- `pop` / `n` — weighted population and observation count
- `decile_val` / `decile_n` — total decile income and decile population

### Scaling up: `f_agg_by_decile_by_sim()`

This function loops over all decile variables × aggregation variables × group
variables for each simulation, then **stacks everything with `bind_rows()`**
adding a `sim` column. This is the pivot point — list becomes one tibble:

```{r step2-aggregate}
agg <- sim_with_deciles |>
  f_agg_by_decile_by_sim(
    var_decile = c("ym___decile"),
    var_agg = c("ym", "dtx_prog1"),
    var_group = c("group_1"),
    wt_var = "hhwt"
  )

dim(agg)
```

```{r step2-structure}
names(agg)
```

```{r step2-show}
agg |>
  select(sim, decile_var, decile, var, level, group_var, group_val) |>
  head(10)
```

Each row is one combination of: **simulation × decile variable × decile ×
aggregated variable × group**. The `sim` column is what enables comparison.

---

## Phase 3: Transform — Incidence and Formatting

### What `f_calc_incidence()` does

This function joins the variable dictionary to get a sign factor (+1 for
benefits/income, −1 for taxes), then computes three measures from each row's
`level` value:

| Measure | Formula | Meaning |
|---------|---------|---------|
| **relative** | `level / decile_val` | Share of this decile's total income |
| **absolute** | `level / total` | Share of all deciles' total for this variable |
| **level** | `level × factor` | Raw weighted sum (sign-adjusted) |

```{r step3-incidence}
incid <- agg |>
  f_calc_incidence(force_abs = TRUE)

incid |>
  select(sim, decile, var, measure, value) |>
  head(9)
```

The data pivots to long format: each original row produces 3 rows (one per
measure type) with `measure` and `value` columns.

### What `f_format_incidence()` does

Applies the column dictionary to produce presentation-quality output:

1. Joins human-readable labels for variables, measures, and groups
2. Converts text columns to ordered factors (preserving logical sort order)
3. Renames columns (`sim` → `Simulation`, `value` → `Value`, etc.)

```{r step4-format}
formatted <- incid |>
  f_format_incidence()

names(formatted)
```

The final structure:

```{r step4-show}
formatted |>
  select(Simulation, `Decile by`, Decile, Variable, Statistics, Value) |>
  head(6)
```

---

## Plotting: The Comparison

The formatted tibble contains all combinations. The app filters to a specific
measure + decile variable + variable, then plots with `sim` as the color:

```{r step5-plot, fig.width=7, fig.height=4}
plot_data <- formatted |>
  filter(
    Statistics == levels(Statistics)[1],
    `Decile by` == levels(`Decile by`)[1],
    Variable == levels(Variable)[1]
  )

f_plot_gg(
  plot_data,
  x_var = "Decile",
  y_var = "Value",
  color_var = "Simulation",
  type = "bar",
  title = "Comparing policies by decile"
)
```

The `Simulation` column distinguishes policies — each gets its own color in the
grouped bar chart.

---

## Complete Pipeline (One Call)
 
In practice, the entire pipeline from raw simulation list to plot-ready data is:

```{r full-pipeline, eval=FALSE}
plot_ready <- dta_sim |>
  f_calc_deciles_by_sim(dec_var = c("ym", "yn"), wt_var = "hhwt", n_dec = 10) |>
  f_agg_by_decile_by_sim(
    var_decile = c("ym___decile", "yn___decile"),
    var_agg = c("ym", "dtx_prog1", "dtx_prog2", "dtr_prog1"),
    var_group = c("group_1", "group_2"),
    wt_var = "hhwt"
  ) |>
  f_calc_incidence(force_abs = TRUE) |>
  f_format_incidence()
```

The Shiny app then progressively filters this dataset by user selections
(decile variable, measure type, grouping, group values) before passing each
subset to `f_plot_gg()`.
